---
title: "EDA"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE, message=FALSE}
library(auk)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(readr)
library(sf)
library(kableExtra)
library(leaflet)

# Set default figure options
knitr::opts_chunk$set(
  fig.width = 6,
  out.width = "90%"
)
```


```{r, include=FALSE, message=FALSE}
observations_selected = read_csv("data/observations_selected_clean.csv")
```

## Season

**Does the season affect the activity of woodcocks?**


```{r}
# Define colors for the seasons
season_colors <- c(
  "Winter" = "#56B4E9", 
  "Spring" = "#009E73",  
  "Summer" = "#F0E442",  
  "Fall" = "#E69F00"
)

# Function to assign colors based on seasons
season_color <- function(season) {
  unname(season_colors[as.character(season)])  # Use color specific to the season
}

# Render the interactive map
leaflet(data = observations_selected) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addCircleMarkers(
    ~longitude, ~latitude,  # Longitude and Latitude
    radius = 3,            # Marker size
    fillColor = ~season_color(season),  # Apply season colors
    color = "grey",        # Grey outline for all markers
    fillOpacity = 0.6,     # Set fill opacity for transparency
    stroke = TRUE,         # Add borders to markers
    weight = 0.5,          # Border thickness
    popup = ~paste("Season:", season, "<br>",
                   "Observation Count:", observation_count, "<br>",
                   "Time:", substr(time_observations_started, 1, 5), "<br>",  # Extract HH:MM
                   "Date:", observation_date)  # Add popup information
  ) %>%
  addLegend(
    position = "bottomright",
    colors = unname(season_colors),  # Colors for the legend
    labels = names(season_colors),  # Labels for the seasons
    title = "Season",
    opacity = 1
  )
```





:::: {style="display: flex;"}

::: {}

```{r, fig.align='center', fig.width=6.5, fig.height=4}
observations_selected <- observations_selected |> 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

season_summary <- observations_selected |> 
  group_by(season) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

ggplot(season_summary, aes(x = season, y = total_observation_count, fill = season)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Winter" = "#56B4E9", 
                               "Spring" = "#009E73", 
                               "Summer" = "#F0E442", 
                               "Fall" = "#E69F00")) +
  labs(
    title = "Observation Count of Woodcocks by Season in NY",
    x = "Season", 
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
       text = element_text(family = "Times", size = 14))
```

<br><br>  

The total observation count in Spring in NY is significantly higher than in any other season.


:::
  
:::{}

```{r, fig.align='center', fig.width=6.5, fig.height=4}
observations_selected <- observations_selected |> 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

season_summary <- observations_selected |> 
  group_by(season) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

# Create a table using kable for display
season_summary |> 
  kable(
    col.names = c("Season", "Total Observation Count"),
    caption = "Table: Total observation count of woodcocks by season in NY"
  )
```

<br><br> 


:::
  
::::
  


<br><br> 







## Time of the day

**When do American woodcocks usually appear?**


```{r}
# map

# Define colors for the times of day
time_colors <- c(
  "Morning" = "#FFD700",    # Gold for Morning
  "Afternoon" = "#FFA500",  # Orange for Afternoon
  "Evening" = "#1E90FF",    # Dodger Blue for Evening
  "Night" = "#4B0082"       # Indigo for Night
)

# Function to assign colors based on time of day
time_color <- function(time_of_day) {
  unname(time_colors[as.character(time_of_day)])  # Use color specific to the time of day
}

# Render the interactive map
leaflet(data = observations_selected) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addCircleMarkers(
    ~longitude, ~latitude,  # Longitude and Latitude
    radius = 3,            # Marker size
    fillColor = ~time_color(time_of_day),  # Apply time of day colors
    color = "grey",        # Grey outline for all markers
    fillOpacity = 0.6,     # Set fill opacity for transparency
    stroke = TRUE,         # Add borders to markers
    weight = 0.5,          # Border thickness
    popup = ~paste("Time of Day:", time_of_day, "<br>",
                   "Observation Count:", observation_count, "<br>",
                   "Time:", substr(time_observations_started, 1, 5), "<br>",  # Extract HH:MM
                   "Date:", observation_date)  # Add popup information
  ) %>%
  addLegend(
    position = "bottomright",
    colors = unname(time_colors),  # Colors for the legend
    labels = names(time_colors),  # Labels for the times of day
    title = "Time of Day",
    opacity = 1
  )
```


:::: {style="display: flex;"}

::: {}


```{r, fig.align='center', fig.width=6.5, fig.height=4}

observations_selected <- observations_selected |> 
  mutate(time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")))

# Summarize observation counts by time of day
time_summary <- observations_selected |> 
  group_by(time_of_day) |> 
  summarise(total_observation_count = n())

# Create the plot
ggplot(time_summary, aes(x = time_of_day, y = total_observation_count, fill = time_of_day)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "Morning" = "#FFD700", 
    "Afternoon" = "#FFA500", 
    "Evening" = "#1E90FF",  
    "Night" = "#4B0082" 
  )) +
  labs(
    title = "Observation Count by Time of Day",
    x = "Time of Day",
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


<br><br>  

The total observation count in Morning.....


:::
  
:::{}


```{r, fig.align='center', fig.width=6.5, fig.height=4}
# table
# Ensure 'time_of_day' is a factor with a specific order
observations_selected <- observations_selected |> 
  mutate(time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")))

# Summarize total observation count by time of day
time_summary <- observations_selected |> 
  group_by(time_of_day) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

# Create a table using kable for display
time_summary |> 
  kable(
    col.names = c("Time of Day", "Total Observation Count"),
    caption = "Table: Total observation count of woodcocks by time of day in NY"
  )
```

<br><br> 


:::
  
::::
  


<br><br> 





Are the difference significant?

```{r}
kruskal_test <- kruskal.test(observation_count ~ time_of_day, data = observations_selected)

# Extract test results
kruskal_result <- data.frame(
  Statistic = round(kruskal_test$statistic, 3),
  Degrees_of_Freedom = kruskal_test$parameter,
  P_Value = format.pval(kruskal_test$p.value, digits = 3, scientific = TRUE)
)

# Save results as a nice table
kruskal_result |> 
  kbl(caption = "Kruskal-Wallis Test Results") |> 
  kable_classic(full_width = FALSE, html_font = "Times")
```

There is strong evidence that the observation counts differ significantly across the time periods (Morning, Afternoon, Evening, Night).


## County

**Observation Counts by County**

:::: {style="display: flex;"}

::: {}

```{r, fig.align='center', fig.width=5, fig.height=4}
# Summarize the data by county
county_summary <- observations_selected |> 
  filter(!is.na(county)) |> 
  group_by(county) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

# Create a bar plot
ggplot(county_summary, aes(x = reorder(county, -total_observation_count), y = total_observation_count, fill = total_observation_count)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Count") +
  labs(
    title = "Observation Counts by County",
    x = "County",
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        legend.position = "right",
        plot.title = element_text(size = 10, hjust = 0.5),  # Smaller title, centered
        axis.title.x = element_text(size = 8),  
        axis.title.y = element_text(size = 8),
        legend.text = element_text(size = 8))
```

:::

:::{}




:::
::::




